Chaotic mixing of tidal debris

### 2015-06-09 ms.tex
very important: added todos

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\msun}{\ensuremath{\mathrm{M}_\odot}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\ident}{\mathbb{1}}
\newcommand{\inttime}{t_{\rm int}}
%\newcommand{\fdrate}{\mathcal{R}_\Omega}

@@

[TODO: take cartoons from talk and include here for description]

@@

We generate isoenergy grids of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass) chosen to span a range of distances comparable to the scale radius of the potential (Section~\ref{sec:potential}), $E=-(397.2~{\rm km~s}^{-1})^2$ (in physical units). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits --- the other major class of orbits in triaxial potentials --- which tend to plunge deep into the inner regions of the potential, however there are appreciable regions of stochastic tube orbits. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the faster disruption expected for stellar systems on radially plunging orbits. The major classes of stable tube orbits circulate about either the major or minor axis. The most numerous orbits in this grid are the short-axis and outer long-axis tubes, but this grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits).

To compute the Lyapunov exponents, we integrate all orbits in the grid for 1000 orbital periods [TODO: re-run for 5000 or 10000?] and use the method described in Section~\ref{sec:lyap}. Figure~\ref{fig:lyapmap} shows the grid of initial conditions (in the $xz$ plane) --- each pixel corresponds to an orbit, and the logarithm of the Lyapunov time (in units of Gyr [TODO: convert to num. periods?]) is mapped to greyscale intensity. The darker pixels have shorter Lyapunov times and are more chaotic. By fixing the integration time (in units of orbital periods), the MLE cannot detect weak chaos and this the majority of orbits appear to be regular because they have exceedingly long Lyapunov times (all white points have $t_\lambda \gtrsim XX$).

For each orbit, we also separately integrate for $\approx$100 orbital periods and use NAFF to compute the fundamental frequencies for the two consecutive sections of 50 orbital periods. Figure~\ref{fig:freqdiff} shows the same grid of initial conditions, but now the greyscale intensity is set by the logarithm of the frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of NAFF --- weak chaos is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs.

@@

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen at random from a single column of the orbit grid based on their frequency diffusion time, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on these three orbits. The initial mass of the clusters is [TODO: 2.5E4 contained within??]. For each simulation, the progenitor is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and evolve for $\approx$16 orbital periods so that each simulation finishes again at subsequent apocenter. Figure~\ref{fig:nbodysims} shows the final particle distributions rotated so that the angular momentum of the progenitor orbit is aligned with the $z$-axis. A thin stream forms on the regular orbit (left column), a more diffuse stream on the mildly chaotic orbit (middle column), and a ``fanned'' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that stream formed on this orbit appears to be more diffuse than the regular orbit stream.

@@

We find that chaotic mixing of small orbit ensembles is detectable over short times, even when the predicted chaotic timescales (from the Lyapunov and frequency diffusion times) are long and therefore that the mean, single-orbit chaos indicators are not well-suited for determining the importance of chaotic diffusion on tidal stream evolution. For example, Figure~\ref{fig:ensemble-density} shows the evolution of the configuration-space density for ensembles evolved around the three orbits (R, W, S) described above. At each time step, we use kernel density estimation (KDE)\footnote{We use an Epanechnikov kernel with a bandwidth set to 1 kpc. We use an implementation from the Python package \texttt{scikit-learn} \citep{scikitlearn}.} with the ensemble of particle positions to estimate the configuration-space density field. We evaluate the density at the positions of each particle, $\rho_i$, and compare to the mean initial density, $\mean{\rho_0}$. Figure~\ref{fig:ensemble-density} shows histograms of the logarithm of $\rho_i/\mean{\rho_0}$ at three different times: after evolution for 4 (left), 8 (middle), and 16 (right) orbital periods. Both the mean density and shape of the density distribution changes over time and the nature of the evolution depends on the strength of chaos along the orbit. The regular orbit distribution stays relatively compact indicating that the density stays fairly uniformly high throughout the stream. For the weakly chaotic orbit, already after 8 orbits the density distribution shows a large tail to smaller densities --- this is ``fanned'' debris --- but maintains a significant number of particles in relatively high densities. The strongly chaotic orbit evolves much faster to low density. Stars with separations much larger than the bandwidth have unresolved density measurements --- this is why there appears to be a pile-up of stars at $\log_{10}\left[\rho_i / \mean{\rho_0}\right] \approx -3$. TODO: do we need to adjust bandwidth as the density evolves?

@@

Motivated by the apparent detection of chaos using these ensemble measures even around the weakly chaotic orbit, we compute the final skewness and mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 16 (parent) orbital periods and store the mean density and skewness at the end of integration for each. Figures~\ref{fig:ensemblemap-meandensity} and \ref{fig:ensemblemap-skewness} again show the grid of initial conditions (Section~\ref{sec:results1}), but now the greyscale indicates the final mean density or skewness, respectively. [TODO: some statement about the structure here, compare to freq. diff time map -- even though freq. diff time are long, still resolve structure in these maps.]

@@

We now repeat the above analysis of the short-time evolution of frequencies, but now for ensembles of orbits around each of the three parent orbits. [TODO: a new metric, measuring the frequency variance, Figure~\ref{fig:freqvar_map} ...]

@@

NAFF recovers the fundamental frequencies for an orbit faster (with a fewer number of terms) when the coordinates used are ``close'' to the angle variables \cite[PL96;][]{papaphilippou96}. PL96 show that a good choice of coordinates for tube orbits are the Poincar\'e symplectic polar coordinates, a set of canonical coordinates similar to cylindrical coordinates. When computing the frequencies for tube orbits, we first align the circulation about the $z$-axis through rotation, transform to Poincar\'e polar coordinates, then use NAFF to measure the fundamental frequencies. We could equivalently use the Cartesian time series, but the convergence of terms is slower (the amplitudes of successive terms decrease slower for Cartesian coordinates). We have tested that our implementation of NAFF returns the same fundamental frequencies in either case for a set of tube orbits. For box orbits, the motion is close to separable in each Cartesian component and we therefore use Cartesian coordinates for estimating the frequencies for these orbits. [TODO: Demo figure of log potential??? convince out implementation is good?]

@@

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/lyap_map.png}
\caption{ A grid of isoenergy orbits initialized on the $xz$ plane. Each pixel in this image represents a single orbit, and the greyscale corresponds to the logarithm of the Lyapunov time (in units of orbital periods). Chaotic orbits with $t_\lambda \gtrsim XX~{\rm T}$ appear regular because the integration time for each orbit is insufficient to resolve weak chaos. [TODO: remove Gyr and express in terms of orbital periods, add title, run Lyapunov grid for 5000 orbital periods instead of 1000?]} \label{fig:lyapmap} 
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/fdiff_map.png}
\caption{The same grid of initial conditions as Figure~\ref{fig:lyapmap}, but now each pixel is colored by the logarithm of the frequency diffusion time (again in units of orbital periods). The frequencies are computed in two consecutive windows, each of which has length equal to 50 orbital periods. The frequencies are measured precisely so that small changes in the frequencies can be detected over just $\approx$10s of orbits. [TODO: remove Gyr and express in terms of orbital periods, add title] } \label{fig:freqdiff}
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/nbody.png}
\caption{Particle positions from the final snapshots of N-body simulations of a globular-cluster-mass satellite disrupting in a triaxial NFW potential (Section~\ref{sec:potential}). Each satellite is evolved for 16 orbital periods, and begins and ends at apocenter. For visualization, the positions have been rotated so that the angular momentum of the progenitor at the final snapshot are aligned with the $z$-axis. Top row shows the positions projected onto the resulting $xy$ plane (the instantaneous orbital plane of the progenitor at the end of the simulation), bottom row shows the positions projected onto the $yz$ plane. From left to right, the columns indicate that the progenitor is on a regular (R), weakly chaotic (W), or strongly chaotic (S) orbit. Even on the weakly chaotic orbit --- where the predicted chaotic timescale is XX orbital periods --- the resulting tidal stream displays stream-fanning and the ``oldest'' debris is more diffuse than in the regular case. [TODO: suptitle 'N-body simulations'?]} \label{fig:nbodysims}
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/ensembles.png}
\caption{Final particle positions after integrating unbound, globular-cluster-sized ensembles of orbits generated around each of the three N-body orbits (Figure~\ref{fig:nbodysims}). The ensembles each contain 1000 particles and begin at pericenter rather than apocenter because they are meant to represent debris disrupted during a single (the first) pericentric passage. The particle positions qualitatively match the morphology of the ``oldest'' debris from the N-body simulations, though there are many more particles in these ensembles than are disrupted during the first pericentric passage of the simulations. [TODO: suptitle 'Orbit ensembles'?] } \label{fig:ensembles}
\end{center}
\end{figure*}

% Figure ??
\clearpage
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/ensemble-density.png}
\caption{ Distributions of the values of the estimated configuration-space density field evaluated at the positions of the ensemble particles after 4 (left), 8 (middle), and 16 (right) orbital periods. Colors represent the three orbits of Figures~\ref{fig:nbodysims} and \ref{fig:ensembles}: the regular, weakly chaotic, and strongly chaotic orbits. The density distribution of the regular orbit ensemble stays compact and decreases slowly. The density of the weakly chaotic ensemble initially evolves quickly to lower density, but then broadens. The strongly chaotic ensemble evolves fast to a state of low density. [TODO: can't distinguish colors when printed...]} \label{fig:ensemble-density}
\end{center}
\end{figure*}

>>>>

% commands
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\msun}{\ensuremath{\mathrm{M}_\odot}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\ident}{\mathbb{1}}
\newcommand{\inttime}{t_{\rm int}}

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\superfreq}{\project{SuperFreq}}

% TO DO
\newcommand{\todo}[2]{{\color{red} TODO: (\MakeUppercase{#1}) #2}}

@@

\todo{APW}{take cartoons from talk and include here for description}

@@

We generate isoenergy grids of initial conditions along the $xz$ ($y=0$) plane\footnote{$x$ is the major and $z$ the minor axis.} with energy (per unit mass) chosen to span a range of distances comparable to the scale radius of the potential ($E=-(397.2~{\rm km~s}^{-1})^2$ in physical units; see Table~\ref{tbl:potential}). We fix $v_x = v_z = 0$, and compute $v_y$ from the energy. The majority of the orbits generated on this grid are tube orbits, which preserve a sense of rotation about one of the principal axes of the potential and are generally centrophobic. Tube orbits are generally less stochastic than box orbits --- the other major class of orbits in triaxial potentials --- which tend to plunge deep into the inner regions of the potential, however there are appreciable regions of stochastic tube orbits. Thin tidal streams are thought to form along tube orbits rather than box orbits because of the faster disruption expected for stellar systems on radially plunging orbits. The major classes of stable tube orbits circulate about either the major or minor axis. This grid generates all of the major orbit classes (short-axis tubes, inner long-axis tubes, outer long-axis tubes, stochastic intermediate-axis, and box orbits), but the most numerous orbits in this grid are the short-axis and outer long-axis tubes.

To compute the Lyapunov exponents, we integrate all orbits in the grid for 1000 orbital periods \todo{apw}{re-run for 5000 or 10000 periods?} and use the method described in Section~\ref{sec:lyap}. Figure~\ref{fig:lyapmap} shows the grid of initial conditions (in the $xz$ plane) --- each pixel corresponds to an orbit, and the logarithm of the Lyapunov time (in units of Gyr \todo{apw}{ convert to num. periods?}) is mapped to greyscale intensity. The darker pixels have shorter Lyapunov times and are more chaotic. By fixing the integration time (in units of orbital periods), the MLE cannot detect weak chaos and the majority of orbits appear to be regular because they have exceedingly long Lyapunov times (all white points have $t_\lambda \gtrsim XX$).

For each orbit, we also separately integrate for $\approx$80 orbital periods and use \superfreq\ to compute the fundamental frequencies for the two consecutive sections of 40 orbital periods. We have found that this is the shortest integration window for which we are able to consistently and accurately recover the frequencies using \superfreq.

 Figure~\ref{fig:freqdiff} shows the same grid of initial conditions, but now the greyscale intensity is set by the logarithm of the frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of NAFF --- weak chaos is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs.

@@

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen at random from a single column of the orbit grid based on their frequency diffusion time, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on these three orbits. The initial mass of the clusters is \todo{kvj}{2.5E4 \msun contained within what again? 35 scale radii? but this is plummer??}. For each simulation, the progenitor is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and evolve for $\approx$16 orbital periods so that each simulation finishes again at subsequent apocenter. Figure~\ref{fig:nbodysims} shows the final particle distributions rotated so that the angular momentum of the progenitor orbit is aligned with the $z$-axis. A thin stream forms on the regular orbit (left column), a more diffuse stream on the mildly chaotic orbit (middle column), and a ``fanned'' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that stream formed on this orbit appears to be more diffuse than the regular orbit stream.

@@

We find that chaotic mixing of small orbit ensembles is detectable over short times, even when the predicted chaotic timescales (from the Lyapunov and frequency diffusion times) are long and therefore that the mean, single-orbit chaos indicators are not well-suited for determining the importance of chaotic diffusion on tidal stream evolution. For example, Figure~\ref{fig:ensemble-density} shows the evolution of the configuration-space density for ensembles evolved around the three orbits (R, W, S) described above. At each time step, we use kernel density estimation (KDE)\footnote{We use an Epanechnikov kernel with a bandwidth set to 1 kpc. We use an implementation from the Python package \texttt{scikit-learn} \citep{scikitlearn}.} with the ensemble of particle positions to estimate the configuration-space density field. We evaluate the density at the positions of each particle, $\rho_i$, and compare to the mean initial density, $\mean{\rho_0}$. Figure~\ref{fig:ensemble-density} shows histograms of the logarithm of $\rho_i/\mean{\rho_0}$ at three different times: after evolution for 4 (left), 8 (middle), and 16 (right) orbital periods. Both the mean density and shape of the density distribution changes over time and the nature of the evolution depends on the strength of chaos along the orbit. The regular orbit distribution stays relatively compact indicating that the density stays fairly uniformly high throughout the stream. For the weakly chaotic orbit, already after 8 orbits the density distribution shows a large tail to smaller densities --- this is ``fanned'' debris --- but maintains a significant number of particles in relatively high densities. The strongly chaotic orbit evolves much faster to low density. Stars with separations much larger than the bandwidth have unresolved density measurements --- this is why there appears to be a pile-up of stars at $\log_{10}\left[\rho_i / \mean{\rho_0}\right] \approx -3$. \todo{apw}{do we need to adjust bandwidth as the density evolves?}

@@

Motivated by the apparent detection of chaos using these ensemble measures even around the weakly chaotic orbit, we compute the final skewness and mean density for orbit ensembles generated around each orbit in the grid described in Section~\ref{sec:results1}. For all ensembles, we integrate the orbits for 16 (parent) orbital periods and store the mean density and skewness at the end of integration for each. Figures~\ref{fig:ensemblemap-meandensity} and \ref{fig:ensemblemap-skewness} again show the grid of initial conditions (Section~\ref{sec:results1}), but now the greyscale indicates the final mean density or skewness, respectively. \todo{apw}{talk about the structure here, compare to freq. diff time map -- even though freq. diff time are long, still resolve structure in these maps.}

@@

We now repeat the above analysis of the short-time evolution of frequencies, but now for ensembles of orbits around each of the three parent orbits. \todo{apw}{a new metric, measuring the frequency variance, Figure~\ref{fig:freqvar_map} ...}

@@

NAFF recovers the fundamental frequencies for an orbit faster (with a fewer number of terms) when the coordinates used are ``close'' to the angle variables \cite[PL96;][]{papaphilippou96}. PL96 show that a good choice of coordinates for tube orbits are the Poincar\'e symplectic polar coordinates, a set of canonical coordinates similar to cylindrical coordinates. When computing the frequencies for tube orbits, we first align the circulation about the $z$-axis through rotation, transform to Poincar\'e polar coordinates, then use NAFF to measure the fundamental frequencies. We could equivalently use the Cartesian time series, but the convergence of terms is slower (the amplitudes of successive terms decrease slower for Cartesian coordinates). We have tested that our implementation of NAFF returns the same fundamental frequencies in either case for a set of tube orbits. For box orbits, the motion is close to separable in each Cartesian component and we therefore use Cartesian coordinates for estimating the frequencies for these orbits. [TODO: Demo figure of log potential??? convince our implementation is good?]

@@

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/lyap_map.png}
\caption{ A grid of isoenergy orbits initialized on the $xz$ plane. Each pixel in this image represents a single orbit, and the greyscale corresponds to the logarithm of the Lyapunov time (in units of orbital periods). Chaotic orbits with $t_\lambda \gtrsim XX~{\rm T}$ appear regular because the integration time for each orbit is insufficient to resolve weak chaos. \todo{apw}{remove Gyr and express in terms of orbital periods, add title, run Lyapunov grid for 5000 orbital periods instead of 1000?}} \label{fig:lyapmap} 
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/fdiff_map.png}
\caption{The same grid of initial conditions as Figure~\ref{fig:lyapmap}, but now each pixel is colored by the logarithm of the frequency diffusion time (again in units of orbital periods). The frequencies are computed in two consecutive windows, each of which has length equal to 50 orbital periods. The frequencies are measured precisely so that small changes in the frequencies can be detected over just $\approx$10s of orbits. \todo{apw}{remove Gyr and express in terms of orbital periods, add title} } \label{fig:freqdiff}
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/nbody.png}
\caption{Particle positions from the final snapshots of N-body simulations of a globular-cluster-mass satellite disrupting in a triaxial NFW potential (Section~\ref{sec:potential}). Each satellite is evolved for 16 orbital periods, and begins and ends at apocenter. For visualization, the positions have been rotated so that the angular momentum of the progenitor at the final snapshot are aligned with the $z$-axis. Top row shows the positions projected onto the resulting $xy$ plane (the instantaneous orbital plane of the progenitor at the end of the simulation), bottom row shows the positions projected onto the $yz$ plane. From left to right, the columns indicate that the progenitor is on a regular (R), weakly chaotic (W), or strongly chaotic (S) orbit. Even on the weakly chaotic orbit --- where the predicted chaotic timescale is XX orbital periods --- the resulting tidal stream displays stream-fanning and the ``oldest'' debris is more diffuse than in the regular case. \todo{apw}{suptitle 'N-body simulations'?}} \label{fig:nbodysims}
\end{center}
\end{figure*}

% Figure ??
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/ensembles.png}
\caption{Final particle positions after integrating unbound, globular-cluster-sized ensembles of orbits generated around each of the three N-body orbits (Figure~\ref{fig:nbodysims}). The ensembles each contain 1000 particles and begin at pericenter rather than apocenter because they are meant to represent debris disrupted during a single (the first) pericentric passage. The particle positions qualitatively match the morphology of the ``oldest'' debris from the N-body simulations, though there are many more particles in these ensembles than are disrupted during the first pericentric passage of the simulations. \todo{apw}{suptitle 'Orbit ensembles'?} } \label{fig:ensembles}
\end{center}
\end{figure*}

% Figure ??
\clearpage
\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/ensemble-density.png}
\caption{ Distributions of the values of the estimated configuration-space density field evaluated at the positions of the ensemble particles after 4 (left), 8 (middle), and 16 (right) orbital periods. Colors represent the three orbits of Figures~\ref{fig:nbodysims} and \ref{fig:ensembles}: the regular, weakly chaotic, and strongly chaotic orbits. The density distribution of the regular orbit ensemble stays compact and decreases slowly. The density of the weakly chaotic ensemble initially evolves quickly to lower density, but then broadens. The strongly chaotic ensemble evolves fast to a state of low density. \todo{apw}{can't distinguish colors when printed...}} \label{fig:ensemble-density}
\end{center}
\end{figure*}
